---
title: "Introduction to openfhe: Fully Homomorphic Encryption in R"
author: "Balasubramanian Narasimhan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to openfhe: Fully Homomorphic Encryption in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What is Fully Homomorphic Encryption?

Fully Homomorphic Encryption (FHE) allows computation on encrypted
data without decryption. A data holder can encrypt sensitive data,
send it to an untrusted server for processing, and receive encrypted
results — all without the server ever seeing the plaintext.

The `openfhe` R package provides an interface to the
[OpenFHE](https://github.com/openfheorg/openfhe-development) C++
library, which implements the major FHE schemes:

- **BFV/BGV**: Exact integer arithmetic
- **CKKS**: Approximate real-number arithmetic
- **FHEW/TFHE**: Boolean circuit evaluation

## Quick Start: Integer Arithmetic (BFV)

```{r bfv-basic}
library(openfhe)

# Create a context for BFV with 128-bit security
cc <- fhe_context("BFV", plaintext_modulus = 65537, multiplicative_depth = 2)

# Generate keys (including evaluation keys for multiplication)
keys <- key_gen(cc, eval_mult = TRUE)

# Encrypt integer vectors
x <- c(1, 2, 3, 4, 5, 6, 7, 8)
ct_x <- encrypt(keys@public, make_packed_plaintext(cc, x), cc = cc)

y <- c(10, 20, 30, 40, 50, 60, 70, 80)
ct_y <- encrypt(keys@public, make_packed_plaintext(cc, y), cc = cc)

# Compute on encrypted data — the server never sees x or y
ct_sum  <- ct_x + ct_y       # element-wise addition
ct_prod <- ct_x * ct_y       # element-wise multiplication

# Decrypt (only the key holder can do this)
result_sum <- decrypt(ct_sum, keys@secret, cc = cc)
set_length(result_sum, 8)
get_packed_value(result_sum)
# [1] 11 22 33 44 55 66 77 88  (exact!)
```

BFV gives **exact** results for integer arithmetic modulo the
plaintext modulus. No approximation error.

## Real-Number Arithmetic (CKKS)

For computations involving real numbers (statistics, ML), CKKS
provides approximate arithmetic:

```{r ckks-basic}
cc <- fhe_context("CKKS",
  multiplicative_depth = 1,
  scaling_mod_size = 50,
  batch_size = 8
)
keys <- key_gen(cc, eval_mult = TRUE)

# Encrypt real-valued vectors
x <- c(0.25, 0.5, 0.75, 1.0, 2.0, 3.0, 4.0, 5.0)
ct <- encrypt(keys@public, make_ckks_packed_plaintext(cc, x), cc = cc)

# Homomorphic operations
ct_doubled <- ct + ct
ct_squared <- ct * ct
ct_scaled  <- ct * 4.0

# Decrypt
result <- decrypt(ct_doubled, keys@secret, cc = cc)
set_length(result, 8)
get_real_packed_value(result)
# [1] 0.5 1.0 1.5 2.0 4.0 6.0 8.0 10.0  (within ~1e-7 tolerance)
```

CKKS results are approximate — the precision depends on the scaling
parameters. For most statistical applications, this is more than
sufficient.

## Boolean Circuits (Binary FHE)

For logic operations on encrypted bits:

```{r binfhe-basic}
ctx <- bin_fhe_context(BinFHEParamSet$STD128, BinFHEMethod$GINX)
sk <- bin_key_gen(ctx)
bin_bt_key_gen(ctx, sk)

# OpenFHE requires independent ciphertext copies for binary gates
ct_and <- eval_bin_gate(ctx, BinGate$AND,
  bin_encrypt(ctx, sk, 1L),
  bin_encrypt(ctx, sk, 0L))
bin_decrypt(ctx, sk, ct_and)  # 0 (FALSE)

ct_or <- eval_bin_gate(ctx, BinGate$OR,
  bin_encrypt(ctx, sk, 1L),
  bin_encrypt(ctx, sk, 0L))
bin_decrypt(ctx, sk, ct_or)   # 1 (TRUE)
```

## Serialization

For distributed protocols, all objects can be serialized:

```{r serial}
tdir <- tempdir()
# Save to files
fhe_serialize(cc, file.path(tdir, "context.bin"))
fhe_serialize(keys@public, file.path(tdir, "pubkey.bin"))
fhe_serialize(ct, file.path(tdir, "ciphertext.bin"))
serialize_eval_keys(file.path(tdir, "mult_keys.bin"), "mult")

# Load (could be in another session)
cc2 <- fhe_deserialize(file.path(tdir, "context.bin"), "CryptoContext")
ct2 <- fhe_deserialize(file.path(tdir, "ciphertext.bin"), "Ciphertext")
```

## Threshold (Multi-Party) Decryption

Multiple parties can jointly decrypt without any single party
holding the complete secret key:

```{r threshold}
cc <- fhe_context("BFV",
  plaintext_modulus = 65537,
  multiplicative_depth = 2,
  features = c(Feature$MULTIPARTY)
)

# Party A generates initial keys
kp_a <- key_gen(cc)

# Party B generates keys using A's public key
kp_b <- multiparty_key_gen(cc, kp_a@public)

# Encrypt with B's public key
ct <- encrypt(kp_b@public, make_packed_plaintext(cc, 1:8), cc = cc)

# Distributed decryption
partial_a <- multiparty_decrypt_lead(cc, kp_a@secret, ct)
partial_b <- multiparty_decrypt_main(cc, kp_b@secret, ct)
result <- multiparty_decrypt_fusion(cc, partial_a, partial_b)
```

## Why openfhe?

The `openfhe` package is the natural upgrade from
[homomorpheR](https://github.com/bnaras/homomorpheR), which
demonstrated privacy-preserving computation using Paillier encryption
(additive homomorphism only). With `openfhe`, R users get:

- **Full homomorphism**: both addition and multiplication on encrypted data
- **Production performance**: backed by the optimized OpenFHE C++ library
- **Multiple schemes**: choose the right tool (exact integers, approximate reals, Boolean logic)
- **Threshold FHE**: distributed trust across multiple parties
- **Serialization**: for federated analytics across institutions

This opens the door to privacy-preserving medical statistics,
federated learning, and secure multi-party computation — directly
from R.

## Further Reading

- [OpenFHE documentation](https://openfhe-development.readthedocs.io/)
- [homomorpheR](https://github.com/bnaras/homomorpheR) — predecessor package
